---
title: "Flux-focused age modeling with fluxcapacitoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flux-focused age modeling with fluxcapacitoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
set.seed(42)
```

## Overview

`fluxcapacitoR` reframes age modeling around *flux* — sedimentation and
accumulation rates — rather than age directly. Starting from an existing
age-depth ensemble (e.g. a Bacon or BChron model stored in a LiPD file), it:

1. estimates sedimentation rate (years per unit depth) for each ensemble
   member,
2. simulates high-frequency annual-layer ("varve") thickness variability with
   tunable persistence (`addVarves()` / `simulateVarves()`), and
3. uses a Metropolis-Hastings search to find the persistence (`ar1`) and
   varve-scaling parameters whose simulated flux series are most consistent
   with the calibrated radiometric date distributions (`ageProbsDT()`).

This vignette runs a deliberately small version of that search on one lake so
it builds quickly; in practice you would use many more iterations.

```{r libs}
library(fluxcapacitoR)
library(lipdR)
library(geoChronR)
library(ggplot2)
library(purrr)
library(dplyr)
```

## Load a dataset

We bundle one example record — Lake Imandra (Holtzman et al., 2024) — in the
package. It carries a chron ensemble (`ageEnsemble`) and the calibrated date
distributions (`distributionTable`) we evaluate against.

```{r load}
L <- readLipd(system.file("extdata", "Imandra.Holtzman.2024.lpd",
                          package = "fluxcapacitoR"))
chron.model <- 1
DT <- L$chronData[[1]]$model[[chron.model]]$distributionTable
```

## Sedimentation rate from the existing ensemble

`getSedRatesFromChronEnsemble()` is a small helper (defined here in the
vignette; a good candidate to migrate into the package) that differences the
age ensemble with depth to get per-member sedimentation rates.

```{r sedrates}
getSedRatesFromChronEnsemble <- function(L, chron.num = 1, model.num = 1,
                                         ens.table.num = 1, depth.var.name = "depth",
                                         age.ens.var.name = "ageEnsemble") {
  et <- L$chronData[[chron.num]]$model[[model.num]]$ensembleTable[[ens.table.num]]
  ae <- et[[age.ens.var.name]]$values
  depths <- et[[depth.var.name]]$values
  srEns <- apply(ae, 2, \(x) diff(x) / diff(depths))
  srDepths <- rowMeans(cbind(depths[-1], depths[-length(depths)]))
  meanSedRate <- apply(ae, 2, \(a, d) diff(range(d)) / diff(range(a)), depths)
  list(ageEns = ae, ageEnsDepths = depths, srDepths = srDepths, srEns = srEns,
       srUnits = paste0(et[[age.ens.var.name]]$units, "/", et[[depth.var.name]]$units),
       depthUnits = et[[depth.var.name]]$units,
       depthResolution = abs(median(diff(srDepths))), meanSedRate = meanSedRate)
}

ensembleData <- getSedRatesFromChronEnsemble(L, model.num = chron.model)
dim(ensembleData$ageEns)
```

We interpolate each member's sedimentation rate onto a regular depth grid and
rebuild ages by cumulative integration, then score the starting ensemble
against the dated distributions with `ageProbsDT()`.

```{r interp}
depth.step <- 0.5
depths.to.model <- seq(min(ensembleData$ageEnsDepths) - depth.step / 2,
                       max(ensembleData$ageEnsDepths) + depth.step / 2, by = depth.step)
depths.to.model.sr <- rowMeans(cbind(depths.to.model[-1], depths.to.model[-length(depths.to.model)]))

interpolated.yrPerDepth <- apply(ensembleData$srEns, 2,
  \(sr) Hmisc::approxExtrap(x = ensembleData$srDepths, y = sr,
                            xout = depths.to.model, method = "linear")$y)
interpolated.yrPerDepth[!is.finite(interpolated.yrPerDepth)] <- 0
interpolated.yrPerDepth[interpolated.yrPerDepth < 0] <- 0

startYears <- apply(ensembleData$ageEns, 2, min, na.rm = TRUE)
sym <- matrix(startYears, ncol = ncol(interpolated.yrPerDepth),
              nrow = nrow(interpolated.yrPerDepth), byrow = TRUE)
interpolated.ageEns <- apply(interpolated.yrPerDepth, 2, \(x) cumsum(x * depth.step)) + sym

origObj <- ageProbsDT(DT, ageEstimates = interpolated.ageEns, depths = depths.to.model.sr)
summary(origObj)
```

## The Metropolis-Hastings search

Priors on the two tuned parameters:

```{r priors}
priorAr1 <- function(ar1) if (ar1 >= 1 | ar1 < 0) 0 else dnorm(ar1, mean = 0.5, sd = 0.25)
priorVarveScalingFactor <- function(v) if (v < 0) 0 else 0.1
```

Search settings. These are tiny so the vignette builds fast — a real run uses
`n.it` in the thousands and `nBest`/`nVarveEns` of ~100+.

```{r mcmc-settings}
n.it <- 40
nBest <- 100
nVarveEns <- 50
ar1Step <- 0.1
varveScalingFactorStep <- 5
```

```{r mcmc}
ar1Chain <- varveScalingFactorChain <- objChain <- matrix(NA, nrow = n.it)
varveScalingFactorChain[1] <- 25
ar1Chain[1] <- 0.5

sorted <- sort(origObj, decreasing = TRUE, index.return = TRUE)
si <- sorted$ix
fracReplaced <- 0
interpolated.ageEns <- interpolated.ageEns[, si[1:nBest]]
interpolated.yrPerDepth <- interpolated.yrPerDepth[, si[1:nBest]]

best <- map(1:nBest, \(s) list(obj = min(sorted$x[1:nBest]),
                               ageEns = interpolated.ageEns[, s],
                               ar1 = NA, varveScalingFactor = NA, i = 0))
objChain[1] <- -Inf

for (i in 2:n.it) {
  ar1Chain[i] <- ar1Chain[i - 1]
  varveScalingFactorChain[i] <- varveScalingFactorChain[i - 1]
  repeat {
    if (i %% 2 == 0) ar1Chain[i] <- ar1Chain[i] + rnorm(1, sd = ar1Step)
    else varveScalingFactorChain[i] <- varveScalingFactorChain[i] + rnorm(1, sd = varveScalingFactorStep)
    priorObj <- log(priorAr1(ar1Chain[i])) + log(priorVarveScalingFactor(varveScalingFactorChain[i]))
    if (is.finite(priorObj)) break
  }

  if (fracReplaced < 1) {
    srEstimate <- interpolated.yrPerDepth[, sample(seq_len(nBest), size = nVarveEns)]
  } else {
    bestEns <- map(best[sample(seq_len(nBest), size = nVarveEns)], "ageEns") |>
      list_c() |> matrix(nrow = length(depths.to.model), ncol = nVarveEns)
    newSrEns <- apply(bestEns, 2, \(x) diff(x) / diff(depths.to.model))
    srEstimate <- apply(newSrEns, 2,
      \(sr) Hmisc::approxExtrap(x = depths.to.model.sr, y = sr,
                                xout = depths.to.model, method = "linear")$y)
    srEstimate[!is.finite(srEstimate)] <- 0
    srEstimate[srEstimate < 0] <- 0
  }

  result <- addVarves(ages = interpolated.ageEns[, sample(seq_len(nVarveEns))],
                      model.depths = depths.to.model.sr,
                      yrPerDepth = srEstimate,
                      totalDepth = diff(range(ensembleData$ageEnsDepths)),
                      varveMean = mean(ensembleData$meanSedRate) * varveScalingFactorChain[i],
                      ar1 = ar1Chain[i], H = NULL,
                      n.varve.ens = nVarveEns, DT = DT, progress = FALSE)

  proposedObj <- mean(sort(result$varvedPriorLogObj, decreasing = TRUE)[1:nBest], na.rm = TRUE) + priorObj
  objChain[i] <- proposedObj
  if (!((objChain[i] - objChain[i - 1]) > log(runif(1)))) {
    objChain[i] <- objChain[i - 1]
    ar1Chain[i] <- ar1Chain[i - 1]
    varveScalingFactorChain[i] <- varveScalingFactorChain[i - 1]
  }

  currentBest <- map_dbl(best, "obj")
  if (any(result$varvedPriorLogObj > min(currentBest))) {
    sorted <- sort(result$varvedPriorLogObj, decreasing = TRUE, index.return = TRUE)
    si <- sorted$ix
    for (s in seq_along(si)) {
      iae <- Hmisc::approxExtrap(result$ageDepths, result$agePriors[, si[s]], xout = depths.to.model)$y
      best[[which.min(currentBest)]] <- list(obj = result$varvedPriorLogObj[si[s]],
                                             ageEns = iae, ar1 = ar1Chain[i],
                                             varveScalingFactor = varveScalingFactorChain[i], i = i)
      currentBest <- map_dbl(best, "obj")
      if (!any(sorted$x[-seq(1, s)] > min(currentBest))) break
    }
  }
  fracReplaced <- mean(!is.na(map_dbl(best, "ar1")))
}

bestObj <- map_dbl(best, "obj")
bestEns <- map(best, "ageEns") |> list_c() |> matrix(nrow = length(depths.to.model), ncol = nBest)
```

## Did the fit improve?

The updated ensemble shifts toward higher log-likelihood against the dated
distributions:

```{r obj-compare}
ggplot() +
  geom_density(aes(x = origObj, fill = "Original")) +
  geom_density(aes(x = bestObj, fill = "Updated"), alpha = 0.5) +
  scale_fill_manual(paste(nBest, "best ensemble members"),
                    values = c("gray30", "red")) +
  labs(x = "log(objective)", y = "Probability density") +
  theme_bw()
```

## Age models, before and after

```{r agemodels, fig.height = 4}
plotModelDistributions(L, model.num = chron.model) |>
  plotTimeseriesEnsLines(X = bestEns, Y = depths.to.model,
                         alp = 0.1, n.ens.plot = 50, color = "red") +
  scale_y_reverse("Depth (cm)") +
  labs(x = "Age (yr BP)", title = paste(L$dataSetName, "updated"))
```

## Sedimentation rate through time

```{r sedrate-plot}
newSrEns <- apply(bestEns, 2, \(x) diff(x) / diff(depths.to.model))
srEstimate <- apply(newSrEns, 2,
  \(sr) Hmisc::approxExtrap(x = depths.to.model.sr, y = sr,
                            xout = depths.to.model, method = "linear")$y)
srEstimate[!is.finite(srEstimate)] <- 0
srEstimate[srEstimate < 0] <- 0

plotTimeseriesEnsRibbons(X = depths.to.model, Y = 1 / interpolated.yrPerDepth) |>
  plotTimeseriesEnsRibbons(X = depths.to.model, Y = 1 / srEstimate, alp = 0.5,
                           color.low = "pink", color.high = "coral", color.line = "red") +
  scale_y_log10("Sedimentation rate (cm/yr)") +
  xlab("Depth (cm)")
```

## Power spectrum of sedimentation rate

We pass `x.lims` explicitly here: the `plotSpectraEns()` in the released
geoChronR errors when `x.lims` is left `NULL` (fixed on the development branch
that this package was split from).

```{r spectra}
spectraEnsOld <- computeSpectraEns(time = depths.to.model, values = 1 / interpolated.yrPerDepth)
spectraEnsNew <- computeSpectraEns(time = depths.to.model.sr, values = 1 / newSrEns)

x.lims <- rev(range(1 / spectraEnsOld$freq))
plotSpectraEns(spec.ens = spectraEnsOld, x.lims = x.lims) +
  labs(x = "Period (cm per cycle)")
```

## Saving results

Everything needed to reconstruct or re-plot a run fits in a single list:

```{r save, eval = FALSE}
keyData <- list(best = best,
                ensembleData = ensembleData,
                depths.to.model = depths.to.model,
                depths.to.model.sr = depths.to.model.sr,
                L = L)
saveRDS(keyData, file = file.path(tempdir(), paste0(L$dataSetName, "-fluxCapacitoR.RDS")))
```
